---
title: 'Case Study 1: How does a bike-share navigate speedy success?'
---

## Introduction

This exploratory analysis case study is a capstone project required for Google Data analytics Professional Certificate. This case study involves a bike-share company's data of its customer's trip details over a period of 12 months (March 2021 - February 2022). The data has been been made available be Motivate International Inc. under this license.

## Scenario:

You are a junior data analyst working in the marketing analyst team at Cyclistic, a bike-share company in Chicago. The director of marketing believes the company's future success depends on maximizing the number of annual memberships. Therefore, your team wants to understand how casual riders and annual members use Cyclistic bikes differently. From these insights, your team will design a new marketing strategy to convert casual riders into annual members. But first, Cyclistic executives must approve your recommendations, so they must be backed up with compelling data insights and professional data visualizations.

## Stakeholders:

-   Lily Moreno: Director of marketing
-   Cyclistic executive team
-   data analytics team

## Objective:

Analyse Cyclistic historical bike trip data to understand how annual members and casual riders use Cyclistic bikes differently.

## Deliverables:

-   Insights on how annual members and casual riders use Cyclistic bikes differently.
-   Provide effective visuals and relevant data to support the insights.
-   Recommendations to convert casual riders into cyclistic members.

## 1. Prepare

#### Data Sources:

A total of 12 data sets have been made available for each month starting from March 2021 to February 2022. Each data set captures the details of every ride logged by the customers of Cyclistic. This data that has been made publicly available and has been scrubbed to omit rider's personal information.

## Setting Up The Environment

### Install required packages

```{r}
library(tidyverse) #helps wrangle data
library(ggplot2)   #helps visualize data
library(lubridate) #helps wrangle date attributes
```

## Setting up the working directory

```{r}
getwd()  #displays your working directory
setwd("C:/Users/Justin/OneDrive/Documents/R Coding/Cyclistic_data/") #sets your working directory to the specified location
```

## Loading Datasets

```{r}
tripdata_202103 <- read.csv("202103-divvy-tripdata.csv") 
tripdata_202104 <- read.csv("202104-divvy-tripdata.csv") 
tripdata_202105 <- read.csv("202105-divvy-tripdata.csv") 
tripdata_202106 <- read.csv("202106-divvy-tripdata.csv") 
tripdata_202107 <- read.csv("202107-divvy-tripdata.csv") 
tripdata_202108 <- read.csv("202108-divvy-tripdata.csv") 
tripdata_202109 <- read.csv("202109-divvy-tripdata.csv") 
tripdata_202110 <- read.csv("202110-divvy-tripdata.csv") 
tripdata_202111 <- read.csv("202111-divvy-tripdata.csv") 
tripdata_202112 <- read.csv("202112-divvy-tripdata.csv") 
tripdata_202201 <- read.csv("202201-divvy-tripdata.csv") 
tripdata_202202 <- read.csv("202202-divvy-tripdata.csv")
```

## Check Column names for each data set for consistency
```{r}
colnames(tripdata_202103)
colnames(tripdata_202104)
colnames(tripdata_202105)
colnames(tripdata_202106)
colnames(tripdata_202107)
colnames(tripdata_202108)
colnames(tripdata_202109)
colnames(tripdata_202110)
colnames(tripdata_202111)
colnames(tripdata_202112)
colnames(tripdata_202201)
colnames(tripdata_202202)
```

## Check data structures and data types for all data frames
```{r}
str(tripdata_202103)
str(tripdata_202104)
str(tripdata_202105)
str(tripdata_202106)
str(tripdata_202107)
str(tripdata_202108)
str(tripdata_202109)
str(tripdata_202110)
str(tripdata_202111)
str(tripdata_202112)
str(tripdata_202201)
str(tripdata_202202)
```

## 2. Data wrangling and Data cleaning
```{r}
all_trips=bind_rows(tripdata_202103,tripdata_202104,tripdata_202105,tripdata_202106,tripdata_202107,tripdata_202108, tripdata_202109,tripdata_202110,tripdata_202111,tripdata_202112,tripdata_202201,tripdata_202202) #Combine all files into a single file
head(all_trips,10)
str(all_trips)
```

## Changing the data type of started_at and ended_at from char to datetime datatype
```{r}
all_trips[['started_at']] <- ymd_hms(all_trips[['started_at']])
all_trips[['ended_at']] <- ymd_hms(all_trips[['ended_at']])
```

## Renaming columns to make them consistent and understandable
```{r}
all_trips <- rename(all_trips,
                     trip_id = ride_id,
                     bike_type = rideable_type,
                     start_time = started_at,
                     end_time = ended_at,
                     from_station_name = start_station_name,
                     from_station_id = start_station_id,
                     to_station_name = end_station_name,
                     to_station_id = end_station_id,
                     usertype = member_casual)
```

## Inspect the new table that has been created
```{r}
colnames(all_trips) #List of column names
nrow(all_trips) #Total no. of rows
dim(all_trips)  #Dimensions of the data frame
head(all_trips,10) #First 10 rows
tail(all_trips,10) #Last 10 rows
str(all_trips)
```

## Add columns that list the date, month, day, and year of each ride
```{r}
all_trips$date <- as.Date(all_trips$start_time)
all_trips$month <- format(as.Date(all_trips$date), '%B')
all_trips$day <- format(as.Date(all_trips$date), "%d")
all_trips$year <- format(as.Date(all_trips$date), "%Y")
all_trips$day_of_week <- format(as.Date(all_trips$date), "%A")
```

## Add a "ride_length" calculation to all_trips (in minutes)
```{r}
all_trips$ride_length <- as.numeric(difftime(all_trips$end_time,all_trips$start_time,units = 'mins'))
```

## Remove "bad" data
The data frame includes a few hundred entries when bikes were taken out of docks and checked for quality by Divvy or ride_length was negative.
We will create a new version of the data frame (v2) since data is being removed
```{r}
all_trips_v2 <- all_trips[!(all_trips$from_station_name == "HQ QR" | all_trips$ride_length<0),]
```

## 3. Descriptive analysis

```{r}
summary(all_trips_v2)  #Statistical summary of data
table(all_trips$usertype) #Checking how many casual riders and members are there
summary(all_trips_v2$ride_length) #Descriptive analysis on ride_length
```

## Compare members and casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype, FUN = mean)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype, FUN = median)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype, FUN = max)
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype, FUN = min)
```

## The days of the week are out of order. Let's fix that.
```{r}
all_trips_v2$day_of_week <- ordered(all_trips_v2$day_of_week, levels=c("Sunday", "Monday", "Tuesday", "Wednesday", "Thursday", "Friday", "Saturday"))
```

## See the average ride time by each day for members vs casual users
```{r}
aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype+all_trips_v2$day_of_week, FUN = mean)
```

## Analyze ridership data by type and weekday
```{r}
all_trips_v2 %>% 
  group_by(usertype, day_of_week)  %>%
  summarise(number_of_rides = n()	 #calculates the number of rides and average duration 
            ,average_duration = mean(ride_length)) 	# calculates the average duration
```

## Number of rides by weekday for each user type
```{r}
all_trips_v2 %>% 
  group_by(usertype, day_of_week) %>%  
  summarise(number_of_rides = n()) %>% 		
  ggplot(aes(x = day_of_week, y = number_of_rides, fill = usertype)) +
  geom_col(position = "dodge") + labs (title = "Total Rides on each day for each usertype")
```

## Average duration by weekday
```{r}
all_trips_v2 %>%
  group_by(usertype, day_of_week) %>%
  summarise(average_duration = mean(ride_length)) %>% 
  ggplot(aes(x = day_of_week, y = average_duration, fill = usertype)) +
  geom_col(position = "dodge") + labs (title = "Average duration on each day for each usertype")
```

## Ordering the months
```{r}
all_trips_v2$month <- ordered(all_trips_v2$month, levels=c("January","February","March","April","May","June","July","August","September","October","November","December"))
```

## Total Rides by months
```{r}
all_trips_v2 %>% 
  group_by(usertype, month) %>% 
  summarise(number_of_rides = n()) %>%
  ggplot(aes(x = month, y = number_of_rides, fill = usertype)) + geom_bar(position = "dodge", stat='identity') +
  labs(
    title = "Total Rides by Month for each user type") +
  theme(legend.position = "bottom") 
```

## Average Duration by months
```{r}
all_trips_v2 %>% 
  group_by(usertype, month) %>% 
  summarise(average_duration = mean(ride_length)) %>%
  ggplot(aes(x = month, y = average_duration, fill = usertype)) + geom_bar(position = "dodge", stat='identity') +
  labs(title = "Average Duration by Month for each user type") +
  theme(legend.position = "bottom") 
```

## Exporting summary file for further analysis
```{r}
File <- aggregate(all_trips_v2$ride_length ~ all_trips_v2$usertype+all_trips_v2$day_of_week, FUN = mean)
write.csv(File, file = "C:/Users/Justin/OneDrive/Documents/R Coding/Cyclistic_data/avg_length.csv")
write.csv(all_trips_v2, file = "C:/Users/Justin/OneDrive/Documents/R Coding/Cyclistic_data/All_trips_data.csv")
